---
title: "Data analysis project - carry out an IPTW causal analysis - Week 4 Graded Quiz"
output: html_notebook
abstract: For this assignment we will use data from Lalonde (1986), that aimed to
  evaluate the impact of  National Supported Work (NSW) Demonstration, which is a
  labor training program, on post-intervention income levels. Interest is in estimating
  the causal effect of this training program on income.\n
---

<br>

```{r message=FALSE, warning=FALSE}
# load pacakges
library(tableone)
library(Matching)
library(ipw) #compute inverse probability of treatment weights
library(survey) #to et weighted estimators
library(ggplot2)

# now load the lalonde data (which is in the MatchIt package)
data(lalonde, package = 'MatchIt')

# fix missing variables issue by creating black and hispan variables separately (using the levels form the race varible of lalonde)
lalonde$black <- as.integer(ifelse(lalonde$race == "black", 1, 0))
lalonde$hispan <- as.integer(ifelse(lalonde$race == "hispan", 1, 0))
# now drop race variable
lalonde <- subset(lalonde, select = -c(race))

View(lalonde)
```

## Dataset information

The data have n=614 subjects and 10 variables.

**age**: age in years. 

**educ**: years of schooling. 

**black**: indicator variable for blacks. 

**hispan**: indicator variable for Hispanics. 

**married**: indicator variable for marital status.

**nodegree**: indicator variable for high school diploma.

**re74**: real earnings in 1974. 

**re75**: real earnings in 1975.

**re78**: real earnings in 1978. 

**treat**: an indicator variable for treatment status.

The **outcome** is re78 – post-intervention income.

The **treatment** is treat – which is equal to 1 if the subject received the labor training and
equal to 0 otherwise.

The **potential confounding** variables are: age, educ, black, hispan, married, nodegree, re74, re75.

### Q1 Tasks: 
#### Fit a propensity score model. Use a logistic regression model, where the outcome is treatment. Include the 8 confounding variables in the model as predictors, with no interaction terms or non-linear terms (such as squared terms). Obtain the propensity score for each subject. Next, obtain the inverse probability of treatment weights for each subject.

```{r}
# 1. fit model
ps_model <- glm(formula = treat~age+educ+black+hispan+married+nodegree+re74+re75,
                data = lalonde,
                family = binomial())
# 2. obtain propensity scores
ps <- ps_model$fitted.values
# 3. weights (inverse probability of true treatment. if true treatment==1: 1/ps, else 1/(1-ps))
wt <- ifelse(test = lalonde$treat==1, yes = 1/ps, no = 1/(1-ps))
# 4. display summary of weights
summary(wt, digits=3)
```

#### Q1 answer: The min. weight is 1.01. The max. weight is 40.01.


### Q2 Tasks: 
#### Find the standardized differences for each confounder on the weighted (pseudo) population. What is the standardized difference for nodegree?

```{r}
# 1. apply weighs to data
lalonde_weighted <- svydesign(ids = ~ 1, data = lalonde, weights = ~ wt) # survey design function
# 2. creat weighted table 1
xvars = colnames(lalonde_weighted)
# filter out treatment and outcome vars
xvars = xvars[colnames(lalonde_weighted) %in% c("treat", "re78") == FALSE]
table1_weighted <- svyCreateTableOne(vars = xvars , strata = "treat", data = lalonde_weighted, test = FALSE) # test = FALSE, to not perform significance testing

# 3. show table with standardized mean differences (SMD)
print(table1_weighted, smd = TRUE) # standardized mean difference (smd)
```

#### Q2 answer: The standardized difference for the confounder 'nondegree' is: 0.112.

### Q3 Tasks: 
#### Using IPTW, find the estimate and 95% confidence interval for the average causal effect. This can be obtained from svyglm.

```{r}
# 1. specify the propensity score model
weightmodel <- ipwpoint(exposure = treat, family = "binomial", link = "logit",
                        denominator = ~ age+educ+black+hispan+married+nodegree+re74+re75,
                        data = lalonde)

# 2. plot of the weights (density plot)
ipwplot(weights = weightmodel$ipw.weights, logscale = FALSE, main = "weights density plot w/o truncation",
        xlim = c(0,41), xlab = "weights") # x-range (0,41) b/c from Q1 we know that min. weight ca. 1 and max weight ca. 40
```


```{r}
# 3. add weights to lalonde dataframe  
lalonde$wt <- weightmodel$ipw.weights

# 4. # fit a marginal structural model (MSM) to compute risk difference
msm <- svyglm(formula = re78 ~ treat, 
               design = svydesign(~1, weights = ~wt, data = lalonde))

# 5. get coefficients
coef(msm)

# 6. compute confidence intervals
cat("The 95% CI for each coefficient:")
confint(msm)
```

#### Q3 answer: The estimand of the coefficient of treatment in the MSM (i.e. average causal effect (ACE)) is: 224.68. The 95% confidence interval (CI) is: (-1559.32, 2008.67).

### Q4 Tasks: 
#### Now truncate the weights at the 1st and 99th percentiles. This can be done with the trunc=0.01 option in svyglm.

```{r}
# 1. specify the propensity score model
weightmodel_trunc <- ipwpoint(exposure = treat, family = "binomial", link = "logit",
                        denominator = ~ age+educ+black+hispan+married+nodegree+re74+re75,
                        data = lalonde,
                        trunc = 0.01) #truncates top and bottom 1% of weights

# add truncated weights to lalonde dataframe
lalonde$wt_trunc <- weightmodel_trunc$weights.trunc

# 2. display summary of truncated weights
summary(lalonde$wt_trunc, digits=3)
```

##### We observe that the truncated weights only range from ca. 1-13 and not anymore more 1-40 as before. Let's have a look at the weights density plot.

```{r}
# 3. plot of the weights (density plot)
ipwplot(weights = weightmodel_trunc$weights.trunc, logscale = FALSE, main = "weights density plot w/ truncation", xlim = c(0,41), xlab = "truncated weights") # x-range (0,41) b/c from Q4 (above) we know that min. weight ca. 1 and max weight ca. 40
```

```{r}
# now we repeat process from Q3:
# 4. # fit a marginal structural model (MSM) to compute risk difference
msm_trunc <- svyglm(formula = re78 ~ treat, 
               design = svydesign(~1, weights = ~wt_trunc, data = lalonde))

# 5. get coefficients
coef(msm_trunc)

# 6. compute confidence intervals
cat("The 95% CI for each coefficient:")
confint(msm_trunc)
```

#### Q4 answer: The estimand of the coefficient of treatment in the MSM (i.e. average causal effect (ACE)) with truncated weights a 1st and 99th percentiles is: 486.93. The 95% confidence interval (CI) is: (-1090.64, 2064.51).