---
title: IV Analysis in R Example - Week 4
output: html_notebook
---

#### Data example: 
##### Using proximity to 4-year college (Z) and the impact it might have on years of schooling (A) and income (Y).
##### Z: instrumental variable, indicator that subject grew up near 4-year college
##### A: treament variable, subject's years of schooling/education
##### Y: outcome variable, subject's income
##### X: covariate variables, e.g. parents' years of education, region of country, age, race, IQ score from test in high-school, etc.

#### Study motivation: 
##### More schooling/education is usually associated with higher income. But what is the actual causal effect of schooling/education? Concern that people who get more schooling are different in other ways (e.g. socioeconomic background, gender, number of years of education of parents, etc.). -> Confounding, both measured and unmeasured, is a concern when answering this causal effect question.
#### Proposal: 
##### Proximity to 4-year college is an instrument (IV) because living near a 4-year college is a type of **encouragement** to go to college and thus get more years of schooling/education.


```{r}
#install instrumental variable package
install.packages("ivpack")
#load package
library(ivpack)
```

```{r}
# read dataset from above loaded package
data("card.data")
cat(names(card.data))
```

#### About the 'card.data' dataset we just loaded:
##### The instrumental variable (IV) is named 'nearc4' (near 4-year college).
##### The outcome variable is named 'lwage' (log of wage)
##### The treatment variable is named 'educ' (number of years of education/schooling)

```{r}
# inspect descriptive dataset summary statistics
mean(card.data$nearc4) # ca. 68% were "encouraged" to live near 4-year college
# histogram of outcome variable (wage) and the treatment variable (educ)
par(mfrow=c(1,2))
hist(card.data$lwage)
hist(card.data$educ) # majority of kids have 12 years of education (high-school degree) and another peak at 16 years for 4-year college degree
```
#### Next: Estimate strength of our IV
##### Estimate strength of the instrumental variable 'nearc4' by measuring the proportion of "compliers". 'Complier' means that a subject that lives close to a 4-year college (i.e. 'nearc4' == 1) should have went to college ('educ12' == 1) because the went to college and end up having more than 12yrs of schooling/education.

```{r}
# first we need to make the 'educ' variable binary by separating subjects above 12yrs education from the rest
educ12 <- (card.data$educ > 12)
card.data$educ12 <- educ12
# estimate proportion of 'compliers' in the subpopulation of 4-year college subjects
propcomp <- mean(card.data$educ12[card.data$nearc4==1]) - mean(card.data$educ12[card.data$nearc4==0]) # E(A|Z=1) = E(A|Z=0) which is the causal effect of encouragement, Z, on treatment received, A.
cat("The proportion of compliers, which also represents the causal effect of the instrumental variable (Z) on treatment received (A), is:", round(propcomp,digits = 4))
# it looks as if living near 4-year college does increase the chances of having more than 12yrs education 
```

#### Next: Estimating the intention to treat effect (ITT effect) and complier average effect (CACE)
##### Note that: Proportion of
```{r}
# first we estimate the intention to treat effect: E(Y|Z=1) - E(Y|Z=0). which is the causal effect of encouragement Z on the outcome Y.
itt <- mean(card.data$lwage[card.data$nearc4==1]) - mean(card.data$lwage[card.data$nearc4==0])
cat("The intention-to-treat effect is:", round(itt,digits = 4), "\n")
cace <- itt/propcomp
cat("The complier averge causal effect (CACE) is:", round(cace, digits = 4), "\n")
# note that the cace is larger than the itt because we make the no-defiers assumption.
```

#### Next: Estimate above causal effects using Two Stage Least Square (2SLS)

##### 2SLS: Step1 - Regress treatment received (A) on the IV/encouragement variable (Z).

```{r}
# fit stage 1 model
s1_model <- lm(formula = "educ12~nearc4" , data = card.data)
# get predicted value of A given Z for each subject
A_hat <- predict.lm(object = s1_model, type = "response")
table(A_hat)
# interpretation: 
# A) encouraged (Z=1) subjects have probability of receiving college education (educ12==1) of 0.54
# B) not encouraged (Z=0) subjects have probability of receiving college education (educ12==1) of 0.42
```

##### 2SLS: Step2 - Regress outcome (Y) on the predicted value of A (A_hat).

```{r}
# fit stage 2 model
card.data$A_hat <- A_hat
s2_model <- lm(formula = "lwage~A_hat", data = card.data)
# the coefficient of the regressor (A_hat) is the complier average causal effect of treatment (A) on the outcome (Y). (compare to same result above)
cat("The averge causal effect of treatment on the outcome (CACE) is:", round(s2_model$coefficients[2], digits = 4), "\n")
```

#### Next: Estimate above causal effects using R package **ivpack** which has built-in Two Stage Least Square (2SLS) functionality.

```{r}
ivmodel <- ivreg(lwage~educ12, ~nearc4, x = TRUE, data = card.data)
table <- robust.se(ivmodel) # Compute robust to heteroskedasticity standard errors for an instrumental variables analysis
table
```

```{r}
cat("The averge causal effect of treatment on the outcome (CACE) is:", round(table[2,1], digits = 4), "\nWith a p-value of", table[2,4], "this result is highly significant. This points to strong evidence of a causal effect of the treatment.")
```

#### Finally: Controlling for covariates (X).
##### Motivation: The IV assumptions might not seem plausible/met. E.g. subjects living near 4-year college live in more expensive houses and thus come from a more socioeconomic stable household (high income families). So we would like to control for these covariates. Ideallly, the instrument Z might meet the IV assumptions after controlling for these covariates.

```{r}
ivmodel2 <- ivreg(lwage ~ educ12 + exper + reg661 + reg662 + reg663 + reg664 + reg665 + reg666 + reg667 + reg668, 
                        ~ nearc4 + exper + reg661 + reg662 + reg663 + reg664 + reg665 + reg666 + reg667 + reg668,
                  x=TRUE, data=card.data)
table2 <- robust.se(ivmodel2) # Compute robust to heteroskedasticity standard errors for an instrumental variables analysis
table2 
```

```{r}
cat("After accounting for covariates which might cause differences between the subjects: The averge causal effect of treatment on the outcome (CACE) is:", round(table[2,1], digits = 4), "\nWith a p-value of", table[2,4], "this result is STILL highly significant. This points to strong evidence of a causal effect of the treatment.")
```

